plyranges: a grammar for transforming genomics data

### 2018-05-10 paper.Rmd
comments and tweaks for the rest of the paper

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved via
the `group_by_overlaps` operator. Continuing with the data from the 
Human Epigenome Roadmap Consortium data, we can count the number of reads
that fall into a fixed bins of size 10000bp over a sorted and indexed 
BAM file of H3K27Me3 methylation marks from the H1 cell line.

We can resuse the annotation information from our BigWig file example, as the
BAM file was aligned to the hg19 genome build. Alignments over
the mitochondrial genome are dropped using the `dropSeqlevels` function
from the Bioconductor package `GenomeInfoDb` @R-gdb.
```{r locations, eval = FALSE}
locations <- bw_file %>%
  get_genome_info() %>% 
  dropSeqlevels("chrM", pruning.mode = "coarse") %>%
  set_genome_info(genome = "hg19") 
```

To generate bins of width 10000bp we can use the `tile` function from 
`GenomicRanges`:

@@

Finally, we can use `group_by_overlaps` with `summarise` to compute the
total number of reads within each window.

```{r, eval = FALSE}
alignments_summary <- bins %>%
  group_by_overlaps(alginments) %>%
  summarise(n_reads = n())
```

@@

We have created a GRanges object from genotyping performed on the
H1 cell line, consisting of approximately two million single nucleotide
polymorphisms (SNP) and short insertion/deletions (indel). The GRanges object consists
of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
log relative intensity of the probes, GC content score over a probe, and the
name of the probe. We can use this information to compute the transition-transversion
ratio, a quality control metric, within each chromosome in GRanges object.

First we filter out any insertion or deletion alleles and 
the SNPs present on the mitochondria then create a logical vector corresponding
to whether there is a transition event.

@@

We can then compute the transition-transversion ratio using the `group_by`
and `summarise` pattern, it is computed as the total number of transition
SNPs divided by the total number of transversion SNPs within each chromosome
(figure \ref{fig:titv-viz}).

```{r}
ti_tv_results <- h1_snp_array %>% 
  group_by(seqnames) %>%
  summarise(n_snps = n(),
            ti_tv = sum(transition) / sum(!transition)) 
```

@@

# Discussion

A caveat to constructing a compatible interface with `dplyr` is that 
`plyranges` makes extensive use of non-standard evaluation in R
(achieved via the `rlang` package @R-rlang). Simply, this
means that computations are performed and evaluated in the context of the 
GRanges objects; emphasising the interactive nature of our API.
Consequently, when programming with `plyranges` a user needs to be aware of
how non-standard evaluation in R works and how to adapt their code 
accordingly. However, with the rise of R packages like `rlang` this process
is becoming less difficult.

While GRanges are an intuitive representation for data measured on genomic regions,
more flexible data structures are required to represent data from multiple sample
experiments. The Bionconductor class SummarizedExperiment is the canonical 
data structure for representing data for combining multiomic measurements 
from multiple samples. The grammar and design of the `plryanges` DSL can be
naturally extended to the SummarizedExperiment.

# Availablilty and Future Work

The `plyranges` package is available on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the `plyranges` package and extend it for use
with more complex data structures  such as the 
SummarizedExperiment class, which can be used for analysing transcriptomic
and variant data. As the `plyranges` interface encourages tidy data 
practices it integrates well with the principles of the grammar of graphics, 
we aim to use it to prepare data for the visualisation of multimodal biological 
datasets.

>>>>

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved
via the `group_by_overlaps` operator.  We bin and count reads from a
H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
Consortium. The bins have a fixed width of 10000bp.

[ML: why not get the seqinfo from the BAM itself?]

We reuse the genome information from our BigWig file, as the
BAM file was aligned to the hg19 genome build. Alignments over
the mitochondrial genome are dropped using the `dropSeqlevels` function
from the Bioconductor package `GenomeInfoDb` @R-gdb.
[ML: seems like we should have a `select_seqlevels()` or something, so
     that we could do `select_seqlevels(-chrM)`.. somehow we have to
     explain to the reader why we are doing all of this seqlevel munging.]
```{r locations, eval = FALSE}
locations <- bw_file %>%
  get_genome_info() %>% 
  dropSeqlevels("chrM", pruning.mode = "coarse") %>%
  set_genome_info(genome = "hg19") 
```

To generate bins of width 10000bp we use the `tile` function from 
`GenomicRanges`:

@@

[ML: seems like a great opportunity to point out the deferred reading]

Finally, we use `group_by_overlaps` with `summarize` to compute the
total number of reads within each window.

[ML: there was a typo in the code: have you actually run this?]

```{r, eval = FALSE}
alignments_summary <- bins %>%
  group_by_overlaps(alignments) %>%
  summarize(n_reads = n())
```

@@

We have created a GRanges object from genotyping performed on the
H1 cell line, consisting of approximately two million single nucleotide
polymorphisms (SNPs) and short insertion/deletions (indels). The GRanges object consists
of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
log relative intensity of the probes, GC content score over a probe, and the
name of the probe. We can use this information to compute the transition-transversion
ratio, a quality control metric, within each chromosome in GRanges object.

First we filter out the indels and mitochondrial variants. Then we
create a logical vector corresponding to whether there is a transition
event.

@@

We compute the per-chromosome transition-transversion ratio using the
`group_by` and `summarize` pattern (figure \ref{fig:titv-viz}).

```{r}
ti_tv_results <- h1_snp_array %>% 
  group_by(seqnames) %>%
  summarize(n_snps = n(),
            ti_tv = sum(transition) / sum(!transition)) 
```

@@

# Discussion and conclusion

[ ML: there should be some quick summary paragraph here ]

A caveat to constructing a compatible interface with `dplyr` is that
`plyranges` makes extensive use of non-standard evaluation in R via
the `rlang` package @R-rlang. Simply, this means that computations are
evaluated in the context of the GRanges objects. Both dplyr and
plyranges are based on the `rlang` language, because it allows for more
expressive code that is free of repeated references to the
container. Implicitly referencing the container is particularly
convenient when programming interactively.  Consequently, when
programming with `plyranges`, a user needs to generally understand
the `rlang` language and how to adapt their code accordingly. Users
familiar with the tidyverse should already have such knowledge.

We aim to continue developing the `plyranges` package and to extend it
for use with more complex data structures, such as the
SummarizedExperiment class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. The
grammar and design of the `plryanges` DSL are naturally extensible to
SummarizedExperiment.

As the `plyranges` interface encourages tidy data practices, it
integrates well with the grammar of graphics [cite ggplot2].  To
achieve responsive performance, interactive graphics rely on lazy data
access and computing patterns, so the deferred mechanisms within
plyranges should help support interactive genomics applications.

The `plyranges` package is available on the Bioconductor project
website [https://bioconductor.org](https://bioconductor.org) or can be
accessed via
Github
[https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
